---
title: "part_2jw_chisqr_liz"
author: "Joe Walderman"
date: "12/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(janitor)
library(broom)
```

## Overview

In Part 2, we'll do an introduction to chi-square test for independence in R using lizard tail breakage data recorded at Jornada Basin Long Term Ecological Research site. 

**Data from:** Lightfoot, D. 2017. Lizard pitfall trap data (LTER-II, LTER-III) ver 36. Environmental Data Initiative. https://doi.org/10.6073/pasta/ff37b4cace16a9943575f3fd7067064e (Accessed 2020-07-23).

The data are in data/lizards.csv of this project. 

## Step 3. Get data & take a look

Read in the lizards.csv data and take a look (using `View(lizards)` in the Console) to familiarize yourself with the data structure: 

```{r}
lizards <- read_csv(here("data", "lizards.csv"))
```
And for fun, let's find the counts of each lizard species in the data: 

```{r}
lizards %>% 
  count(spp) %>% 
  arrange(-n) # negative means arrange in decreasing order
```


The most commonly observed lizard here (UTST) is *Uta stansburiana*, or commonly the side-blotched lizard. We'll continue by exploring just that lizard. 

## Make a subset

Make a subset from `lizards` that only contains observations for: 
 
- side-blotched lizards (UTST)
- at sites "IBPE" (enclosure grassland site), "RABB" (rabbit mesquite site), and "EAST" (east tarbush site)
- where the tail condition is recorded as either "W" (whole) or "B" (broken)

```{r}
utst <- lizards %>% 
  filter(spp == "UTST") %>% 
  filter(site %in% c("IBPE", "RABB", "EAST")) %>%
  filter(tail %in% c("B", "W"))
```

## Find counts and proportions of lizards with broken or whole tails at each site


```{r}
utst %>% 
  count(site, tail)
```


The `janitor::tabyl()` function gets this pretty close to contingency table format

```{r}
tail_counts <-utst %>% 
  janitor::tabyl(site, tail)

tail_proportions <- tail_counts %>% 
  adorn_percentages() %>% # showing percentage of lizards at each site with broken or whole tails
  adorn_pct_formatting(digits = 2) %>%  # rounding decimals
  adorn_ns() #actual counts in each cell as well in parenthesis
```


## Make contingency table and run chi-square

First, we see that the site is its own variable - when we run chi-square, we really only want the counts in the table. 

We can use `column_to_rownames` to convert entries in a column to stored rownames 

```{r}
lizard_ct <- tail_counts %>% #pull column information into rownames
  column_to_rownames(var = "site") # now we just have two columns and ready to be fed into chi sqrd
```

## Chi-square

**Now** this is in shape to run chi-square. 

What are we asking? Are site and tail breakage independent? In other words: is there a significant association between site and tail breakage? 

```{r}
lizard_x2 <- chisq.test(lizard_ct)
lizard_x2

lizard_tidy <- broom::tidy(lizard_x2)
```
X-square value is an aggregate value indicating how different the values in our table are to what we would expect in null hypothesis scenario that site and tail breakage are not independent. *Degrees of freedom are the number of rows minus one time the number of columns minus one*

There is a significant association (non-independence) between site and lizard tail breakage ($\chi$^2^(`r lizard_tidy$parameter`) = `r lizard_tidy$statistic`, *p* = `r lizard_tidy$p.value`).

Remember: you'd also want to report the actual counts and proportions in each group! 
